### 2021-09-29 papers/modeling-of-combined-lfr-csp-sco2-paper/content.tex
edits to summary and conclusion

Recompression supercritical CO$_2$ Brayton cycles are promising cycle configurations offering higher efficiencies, compact design, and reduced turbomachinery cost while operating with non toxic working fluid. High efficiencies, in some simple recompression cycles reaching 50\% efficiency \cite{turchi_2013,wright_2009}, allows these cycles to be a competitive alternative to steam Rankine and air Brayton cycles over a range of temperatures.
\bl{Updated above as sCO2 can also be used for Coal - it is a cycle vs cycle argument rather than a heat source vs heat source argument -BW}
Various sCO$_2$ Brayton cycles have been modeled with the recompression cycle having efficiency advantages over other proposed cycle arrangements \cite{turchi_2013,ahn_2014,wang_2018}.  Due to the benefits of sCO$_2$ Brayton cycles, the United States Department of Energy is investigating these conversion cycles for use with heat sources including nuclear and solar \cite{doe_2012}. Multiple project funding opportunities are established with the National Energy Technology Laboratory offering 144 million dollar reward for demonstration and performance verification of a sCO$_2$ Brayton cycle \cite{netl_2016} and the Office of Energy Efficiency \& Renewable Energy offering 2.6 million dollar reward in their Brayton Energy project for a integrated CSP receiver, TES, and power block in one sCO$_2$ Brayton system \cite{seto_2015}. Utilizing complementary technologies, specifically solar concentrating power and lead-cooled fast reactors, can offset the drawbacks of each. Coupling these technologies into an interconnected cycle allows for consistent generation, independent of weather or time of day, and thermal storage for high dispatchability during high grid demand periods.

\bl{not a big issue but do we want to describe as sCO2 or sCO2 Brayton? In project nomenclature to date I have used sCO2 with the logic that sCO2 'isnt really a Brayton cycle' but both are common -BW}
A CSP has an array of mirrors concentrating solar rays towards a receiver to generate thermal energy therefore causing a dependency on weather conditions and time of day. Previous research has modeled CSP technology as the thermal source for Brayton sCO$_2$ cycles with promising results in efficiency gains and high temperature thermal energy storage \cite{turchi_2013, iverson_2013, ho_2015, wang_2018, neises_2020}. Thermal energy gathered from the CSP is stored in thermal energy storage which is dispatched into the sCO$_2$ Brayton cycle when grid demand increases. With a comparable temperature to the hot TES, 560 $^{\circ}$C, a LFR is capable of transferring heat to thermal storage for later dispatch. An LFR uses fission reactions to heat a heat transfer fluid, in this case liquid lead, to temperatures of 650 $^{\circ}$C. Nuclear sCO$_2$ Brayton cycles have been studied with similar gains in efficiency seen in CSP sCO$_2$ Brayton cycles \cite{ dostal_2004, luo_2020}. Physical testing of simple sCO$_2$ Brayton cycles is researched at Sandia National Labs in Arvada, Colorado, with break-even conditions, where the turbine produces enough power to offset compressor demand, being achieved \cite{wright_2011}. In addition to break-even conditions, research from the Supercritical CO$_2$ Brayton Cycle Integral Experiment Loop, in Korea Atomic Energy Institute, achieved electrical generation in a simple sCO$_2$ Brayton cycle \cite{cha_2016}.

\bl{I would bullet out the next three citations-BW}
Various studies on complimentary solar-nuclear systems have been accomplished with several reviewed as follows:

@@

\bl{I think the last citation merits its own paragraph and a little further discussion (1-2 more sentences). You could start by saying that it is prior work on the LFR-CSP-sCO2 concept. Then describe its cycle configuration with the high temperature salt. Then in the following sentence (the stuff you currnetly have highlighted) you can say that your work is investigating higher lead temperatures than Wang (because of Westinghouse LFR) and lower temperature salt (technology readiness) and this leads to very different cycle configurations and options than Wang, as will be discussed. The overall tone is that you work recognizes and builds on Wang, rather than needing to justify its existance despite Wang - BW}

@@

\mw{How do we know that 400C is the optimal temperature? This information is given to us by Westinghouse, right? -BW} 
\bl{In an LFR, the outlet temperature is limited by materials considerations. The flow velocity is limited by erosion of the fuel (if the lead gets too fast, it erodes the fuel). We want the core to be compact. Therefore the mass flow is limited. This means that a lower inlet temperature gives a higher power (subject to constraints on fuel temperautre, which is also a function of power density). There is a limit of about 340C, below which the lead might freeze (which is operationally unacceptable), but until that value, lower inlet = higher core power. However, from thermodynamics, higher inlet = higher efficiency. There is a trade-off between driving the core harder and getting better cycle performance. The trade-off gives 400C as a temperature that doesnt overly compromise core power while still providing reasonable efficiency. Raising this temperature higher would drop the core power below the 950 MW -BW}

@@

\bl{I dont think the above covers the case where the salt discharges. It might be worth eliciting how the operating modes change as power demand increases. From highest to lowest demand. In all modes CSP will charge the salt is sun is available: (1) LFR and S2C both dispatch, maximum power (2) As power demand decreases, S2C ramps down until no power dispatched through salt (i.e. salt is charged) (3) LFR dispatches, no dispatch from S2C (4) LFR ramps down and gradually transfers greater and greater fraction of power to salt using C2S. (5) No electricity dispatched, all LFR heat passes through C2S -BW
}

@@

Two separate sensitivity studies on the LFR inlet temperature are completed for C-LFR-ON. The constrained study is calculated by setting the LFR inlet temperature to the design value of $400^{\circ}$C, which is a requirement of the LFR primary circuit to maximize power output within material limits. In addition to the constrained studies, unconstrained studies are required to test the penalties that LFR inlet temperature has on efficiency. The unconstrained study is performed by gradually increasing the mass flow to the main compressor through a parametric study while maximizing cycle efficiency. In Figure \ref{c-lfr-on}, the location of the C2S heat exchanger while charging falls between state point 5 and 6 in any of the studied charging configurations: parallel, pre, circulator or post. 
\bl{flip the order of this para: constrained then unconstrained. State that the unconstrained is a 'sensitivty' case to determine the penalty of the constraint-BW}

@@

The identical inlet temperatures due to the parallel configuration makes the C-1HTR1T-ON cycle configuration restricted. Another single cycle configuration is desired to allow dissimilar inlet temperatures for the CSP HX and LFR HX while additionally testing the effect that mixing flows downstream from the HTR has on cycle efficiency. This cycle, C-2HTR3T-ON, can be seen in Figure \ref{c-2htr3t-on} and has two high temperature recuperators and three turbines. The LFR is powering one turbine, T1, and transferring unused heat to the flow entering LFR HX through a dedicated high temperature recuperator, HTR. The cycle with heat addition from the CSP also contains two separate turbines, diagrammatically shown as one turbine T2, while having a dedicated high temperature recuperator, HTR2. The two turbines displayed as T2 can be modeled as a singular turbine because their isentropic efficiencies are identical causing the inlet and outlet conditions of the turbines to be consistent with a singular turbine. Additionally, the power produced by the two turbines is proportional to mass flow rate, each receives a fraction of the mass flow rate therefore producing the same fraction of power. Summing these power fractions together yields the total power of a singular turbine in the same position. 
\mw{Need to explain how T2 is actually 2 separate turbines represented diagramatically as 1, and why we can model it this way.-BW} 
After the high temperature recuperators, the two flows are combined and sent to the LTR hot side. 
\bl{this is more about having a compromise option of intermediate complexity and efficiency}
\bl{also why this isnt really an addition turbine, it is using the turbines that power the compressors -BW}

@@

Combining the two-cycle efficiencies using Equation \ref{eq-eta-2cycle} gives a more direct comparison to the single cycle configurations; C-1HTR1T and C-2HTR3T. Two combined efficiencies are calculated for the two-cycle configuration combination. The first has the highest efficiency, combining the unconstrained C-LFR-ON LFR low temperature and C-CSP-ON $440^{\circ}$C cold TES cases. This combined efficiency yields a value of 46.05\%. The second combined configuration has a lower efficiency with favorable LFR characteristics and solar salt mass flow rate. This combination has the constrained C-LFR-ON LFR low temperature and C-CSP-ON $390^{\circ}$C cold TES cases with a combined efficiency of 45.05\%. 
\bl{can you produce a combined efficiency which is total cycle electrical output/total heat input for C-LFR-ON + C-CSP-ON in combination-BW}

@@

C-LFR-CIRC is a thermal energy storage charging technique which does not suffer from significant losses associated with turbines, compressors, and heat exchangers. The separate recirculation cycle is modeled extracting heat directly from the LFR, passes the high temperature mass flow through a recirculating pump, exchanges the heat into the hot TES through C2S, then removes heat in a chiller to bring the temperature of the flow down to the inlet temperature of the LFR, ideally $400^{\circ}$C. This is the most adoptable TES charging technique with a wide range of possible TES charging temperatures while transferring a majority of the heat from the LFR into the hot TES. Multiple temperatures for the cold TES are studied; $390^{\circ}$C, $410^{\circ}$C, and $440^{\circ}$C. Of these three temperatures, the highest efficiency is a cold TES temperature of $390^{\circ}$ with a heat storage efficiency of 100\%. The heat storage efficiency reduces as the temperature of the cold TES is increased; the $410^{\circ}$C case has an efficiency of 90\% while the $440^{\circ}$C case has an efficiency of 74\%. This reduction in efficiency is due to the inlet temperature condition on the LFR, the higher the temperature of the cold TES the higher the temperature of the C2S outlet and therefore more heat is being extracted through the chiller to bring the flow temperature down to the LFR inlet temperature. The non-charging cycle configuration that this charging cycle is implemented in requires further consideration. For C-LFR-CIRC, the highest efficiencies are achieved when the cold TES temperature is $10^{\circ}$C less than the inlet to the LFR, this is due to the approach temperature on the C2S heat exchanger. If the desired temperature of cold TES is $390^{\circ}$C, the non-charging cycle configurations with sCO$_2$ LFR inlet temperatures of $400^{\circ}$C are required for highest efficiency. A drawback of this TES charging technique is the additional components needed for the circulation loop, with an additional chiller and circulator component, adding cost. 
\bl{need to comment on how this interacts with the choice of cycle configuration. If the LFR temperature is 400C and cold TES is 390C then CSP inlet must be 380C and this requires 2HTR3T or separate cycles. If 1HTR1T is used, then 410C+ is needed which implies an efficiency penalty -BW}

@@

This paper analyzed contending sCO$_2$ Brayton cycle configurations for complimentary CSP and LFR cycles with non-charging cycle configurations focusing on electrical generation. An additional operating mode, hot TES charging, is considered where the CSP hot TES is charged when grid demand is low. This operating mode positions the heat extraction point at different locations around the turbine in a sCO$_2$ Brayton recompression cycle. The performance parameters that are used for comparison of the cycles are cycle efficiency when the cycle is producing electricity and heat storage efficiency when the focus is on energy storage for later use in the charging mode. These efficiencies are maximized using flow splitter values at different values of cold TES temperature and constrained or unconstrained sCO$_2$ inlet LFR temperature. The conclusions drawn in this paper are as follows:

\textit{Non-charging configurations:}
\begin{itemize}
    \item	Two-cycle configuration C-LFR-ON and C-CSP-ON: Offers the highest cycle efficiency because of the freedom of the two cycles to operate independently when generating electrical power. The cycles individually are simple recompression but lead to redundancy in cycle components increasing cost and complexity.
    \item	C-1HTR1T-ON: Single cycle configuration that is the least complex, therefore cost efficient and compact, with heat additions from the CSP and LFR in parallel. This cycle has the lowest cycle efficiencies with constrained inlet temperatures to the LFR and CSP, limiting versatility.  
    \item   C-2HTR3T-ON: Single cycle configuration that has dedicated turbines and HTRs for the LFR and CSP. This cycle makes a compromise between complexity with reduced cycle component redundancy, flexibility for dissimilar inlet temperatures on the CSP and LFR inlet, and median efficiency. 
\end{itemize}

\textit{Thermal energy storage charging techniques:}
\begin{itemize}
    \item	C-LFR-PRE: Charging technique with the turbine prior to C2S heat exchanger with a valve that bypasses the turbine, demands a high outlet temperature for the LFR to be effective. Due to LFR outlet temperature limitations, this configuration has the lowest heat storage efficiency. 
    \item	C-LFR-PAR: Charging technique that splits the flow directly after the LFR with the turbine and C2S heat exchanger in parallel. The heat storage efficiency is median of the tested charging techniques with a lower cold TES temperature having no effect other than increasing the solar salt mass flow rate increasing CSP component sizes.
    \item   C-LFR-CIRC: Flexible technique with a heat storage efficiency being highly dependent on cold TES temperature and LFR inlet temperature. Able to achieve heat storage efficiency to near 100\% by eliminating losses associated with the turbines and compressors. The additional chiller and circulator components in the dedicated circulation loop increase complexity and cost. 
\end{itemize}

>>>>

Recompression supercritical CO$_2$ Brayton cycles are promising cycle configurations offering higher efficiencies, compact design, and reduced turbomachinery cost while operating with non toxic working fluid. High efficiencies, in some simple recompression cycles reaching 50\% efficiency \cite{turchi_2013,wright_2009}, allows these cycles to be a competitive alternative to steam Rankine and air Brayton cycles over a range of temperatures. Various sCO$_2$ Brayton cycles have been modeled with the recompression cycle having efficiency advantages over other proposed cycle arrangements \cite{turchi_2013,ahn_2014,wang_2018}.  Due to the benefits of sCO$_2$ Brayton cycles, the United States Department of Energy is investigating these conversion cycles for use with heat sources including nuclear and solar \cite{doe_2012}. Multiple project funding opportunities are established with the National Energy Technology Laboratory offering 144 million dollar reward for demonstration and performance verification of a sCO$_2$ Brayton cycle \cite{netl_2016} and the Office of Energy Efficiency \& Renewable Energy offering 2.6 million dollar reward in their Brayton Energy project for a integrated CSP receiver, TES, and power block in one sCO$_2$ Brayton system \cite{seto_2015}. Utilizing complementary technologies, specifically solar concentrating power and lead-cooled fast reactors, can offset the drawbacks of each. Coupling these technologies into an interconnected cycle allows for consistent generation, independent of weather or time of day, and thermal storage for high dispatchability during high grid demand periods.
A CSP has an array of mirrors concentrating solar rays towards a receiver to generate thermal energy therefore causing a dependency on weather conditions and time of day. Previous research has modeled CSP technology as the thermal source for Brayton sCO$_2$ cycles with promising results in efficiency gains and high temperature thermal energy storage \cite{turchi_2013, iverson_2013, ho_2015, wang_2018, neises_2020}. Thermal energy gathered from the CSP is stored in thermal energy storage which is dispatched into the sCO$_2$ Brayton cycle when grid demand increases. With a comparable temperature to the hot TES, 560 $^{\circ}$C, a LFR is capable of transferring heat to thermal storage for later dispatch. An LFR uses fission reactions to heat a heat transfer fluid, in this case liquid lead, to temperatures of 650 $^{\circ}$C. Nuclear sCO$_2$ Brayton cycles have been studied with similar gains in efficiency seen in CSP sCO$_2$ Brayton cycles \cite{ dostal_2004, luo_2020}. Physical testing of simple sCO$_2$ Brayton cycles is researched at Sandia National Labs in Arvada, Colorado, with break-even conditions, where the turbine produces enough power to offset compressor demand, being achieved \cite{wright_2011}. In addition to break-even conditions, research from the Supercritical CO$_2$ Brayton Cycle Integral Experiment Loop, in Korea Atomic Energy Institute, achieved electrical generation in a simple sCO$_2$ Brayton cycle \cite{cha_2016}.

Various studies on complimentary solar-nuclear systems have been accomplished with several reviewed as follows:

@@

@@

\mw{How do we know that 400C is the optimal temperature? This information is given to us by Westinghouse, right? -BW}

@@

@@

Two separate sensitivity studies on the LFR inlet temperature are completed for C-LFR-ON. The constrained study is calculated by setting the LFR inlet temperature to the design value of $400^{\circ}$C, which is a requirement of the LFR primary circuit to maximize power output within material limits. In addition to the constrained studies, unconstrained studies are required to test the penalties that LFR inlet temperature has on efficiency. The unconstrained study is performed by gradually increasing the mass flow to the main compressor through a parametric study while maximizing cycle efficiency. In Figure \ref{c-lfr-on}, the location of the C2S heat exchanger while charging falls between state point 5 and 6 in any of the studied charging configurations: parallel, pre, circulator or post.

@@

The identical inlet temperatures due to the parallel configuration makes the C-1HTR1T-ON cycle configuration restricted. Another single cycle configuration is desired to allow dissimilar inlet temperatures for the CSP HX and LFR HX while additionally testing the effect that mixing flows downstream from the HTR has on cycle efficiency. This cycle, C-2HTR3T-ON, can be seen in Figure \ref{c-2htr3t-on} and has two high temperature recuperators and three turbines. The LFR is powering one turbine, T1, and transferring unused heat to the flow entering LFR HX through a dedicated high temperature recuperator, HTR. The cycle with heat addition from the CSP also contains two separate turbines, diagrammatically shown as one turbine T2, while having a dedicated high temperature recuperator, HTR2. The two turbines displayed as T2 can be modeled as a singular turbine because their isentropic efficiencies are identical causing the inlet and outlet conditions of the turbines to be consistent with a singular turbine. Additionally, the power produced by the two turbines is proportional to mass flow rate, each receives a fraction of the mass flow rate therefore producing the same fraction of power. Summing these power fractions together yields the total power of a singular turbine in the same position. 
\mw{Need to explain how T2 is actually 2 separate turbines represented diagramatically as 1, and why we can model it this way.-BW} 
After the high temperature recuperators, the two flows are combined and sent to the LTR hot side.

In practice, sCO$_2$ cycles typically have 3 turbines, with 2 of these driving the compressor and recompressor. Therefore, this configuration will not in general require additional turbines compared to the C-1HTR1T-ON configuration. Furthermore, it is anticipated that the cost of the high temperature recuperator is likely most related to its volume, and hence having two smaller high temperature recuperators is unlikely to cost significantly more than one large one.

@@

Combining the two-cycle efficiencies using Equation \ref{eq-eta-2cycle} gives a more direct comparison to the single cycle configurations; C-1HTR1T and C-2HTR3T. Two combined efficiencies are calculated for the two-cycle configuration combination. The first has the highest efficiency, combining the unconstrained C-LFR-ON LFR low temperature and C-CSP-ON $440^{\circ}$C cold TES cases. This combined efficiency yields a value of 46.05\%. The second combined configuration has a lower efficiency with favorable LFR characteristics and solar salt mass flow rate. This combination has the constrained C-LFR-ON LFR low temperature and C-CSP-ON $390^{\circ}$C cold TES cases with a combined efficiency of 45.05\%.

@@

C-LFR-CIRC is a thermal energy storage charging technique which does not suffer from significant losses associated with turbines, compressors, and heat exchangers. The separate recirculation cycle is modeled extracting heat directly from the LFR, passes the high temperature mass flow through a circulator, exchanges the heat into the hot TES through C2S, then removes heat in a chiller to bring the temperature of the flow down to the inlet temperature of the LFR, ideally $400^{\circ}$C. This is the most adoptable TES charging technique with a wide range of possible TES charging temperatures while transferring a majority of the heat from the LFR into the hot TES. Multiple temperatures for the cold TES are studied; $390^{\circ}$C, $410^{\circ}$C, and $440^{\circ}$C. Of these three temperatures, the highest efficiency is a cold TES temperature of $390^{\circ}$ with a heat storage efficiency of 100\% (not accounting for pressure drop losses, which will be small). The heat storage efficiency reduces as the temperature of the cold TES is increased; the $410^{\circ}$C case has an efficiency of 90\% while the $440^{\circ}$C case has an efficiency of 74\%. This reduction in efficiency is due to the inlet temperature condition on the LFR, the higher the temperature of the cold TES the higher the temperature of the C2S outlet and therefore more heat is being extracted through the chiller to bring the flow temperature down to the LFR inlet temperature.

There is an interaction between the temperatures selected here and the non-charging cycle configuration. For C-LFR-CIRC, the highest efficiencies are achieved when the cold TES temperature is $10^{\circ}$C less than the inlet to the LFR, this is due to the approach temperature on the C2S heat exchanger. If the sCO$_2$ LFR inlet temperature is constrained to $400^{\circ}$C, then the desired cold TES temperature is $390^{\circ}$C to allow for the apporach temperature. If the cold TES is at a higher temperature, then the chiller is needed. This then implies that the sCO$_2$ inlet temperature in the salt-to-sCO2 heat exchanger is  $380^{\circ}$C. From the results of the previous section, this is clearly achievable for the C-2HTR3T-ON configuration and the separate cycle configuration, but for the C-1HTR1T-ON configuration the LFR and CSP inlet temperatures are constrained to be the same. In this case, the LFR and CSP inlet temperatures would then be $400^{\circ}$C, implying a cold TES temperature of  $410^{\circ}$C and hence a 90\% heat storage efficiency with a requirement for a chiller.

@@

This paper analyzed contending sCO$_2$ Brayton cycle configurations for complimentary CSP and LFR cycles with non-charging cycle configurations focusing on electrical generation. An additional operating mode, hot TES charging, is considered where the CSP hot TES is charged by the LFR when grid demand is low. This operating mode positions the heat extraction point at different locations around the turbine in a sCO$_2$ Brayton recompression cycle. The performance parameters that are used for comparison of the cycles are cycle efficiency when the cycle is producing electricity and heat storage efficiency when the focus is on energy storage for later use in the charging mode. These efficiencies are maximized using flow splitter values at different values of cold TES temperature and constrained or unconstrained sCO$_2$ inlet LFR temperature. The conclusions drawn in this paper are as follows:

\textit{Non-charging configurations:}
\begin{itemize}
    \item	Two-cycle configuration C-LFR-ON and C-CSP-ON: Offers the highest cycle efficiency because of the freedom of the two cycles to operate independently when generating electrical power. The cycles individually are simple recompression but there is a lack of component synergies.
    \item	C-1HTR1T-ON: Single cycle configuration that is the least complex, therefore cost efficient and compact, with heat additions from the CSP and LFR in parallel. With both CSP and LFR discharging, combined cycle efficiency is reduced by ~0.8\% compared to the two cycle configuration. 
    \item   C-2HTR3T-ON: Single cycle configuration that has dedicated turbines and HTRs for the LFR and CSP. This cycle makes a compromise between complexity with reduced cycle component redundancy, flexibility for dissimilar inlet temperatures on the CSP and LFR inlet, and median efficiency. The efficiency is only marginally higher than C-1HTR1T-ON, but heat storage efficiency when using the LFR to charge the salt is improved due to flexibility on setting the CSP and LFR inlet temperatures. As the additional turbines are likely present in practice anyway, this cycle is not anticipated to significantly increase cost and complexity over C-1HTR1T-ON. The relative merits of C-2HTR3T-ON and C-1HTR1T-ON are therefore similar.
\end{itemize}

\textit{Thermal energy storage charging techniques:}
\begin{itemize}
    \item    C-LFR-POST: Charging technique with the turbine subsequent to C2S heat exchanger. Infeasible due to lower thermodynamic efficiency. 
    \item	C-LFR-PRE: Charging technique with the turbine prior to C2S heat exchanger with a valve that bypasses the turbine, demands a high outlet temperature for the LFR to be effective. Due to LFR outlet temperature limitations, this configuration has the next lowest heat storage efficiency.
    \item	C-LFR-PAR: Charging technique that splits the flow directly after the LFR with the turbine and C2S heat exchanger in parallel. The heat storage efficiency is higher than C-LFR-PRE.
    \item   C-LFR-CIRC: Flexible technique with a heat storage efficiency being highly dependent on cold TES temperature and LFR inlet temperature. Able to achieve heat storage efficiency to near 100\% by eliminating losses associated with the turbines and compressors. For C-1HTR1T-ON, the heat storage efficiency is limit to 90\% because the inlet temperatures to the LFR and CSP are constrained to be the same. The additional chiller and circulator components in the dedicated circulation loop increase complexity and cost. 
\end{itemize}

C-LFR-PAR and C-LFR-CIRC are therefore options going forward, with a trade-off between heat storage efficiency and component costs, with the performance of C-LFR-CIRC also depending on the choise of charging configuration.
